---
title: "NYPD Police Shootings Analysis"
author: "------"
date: "2026-09-18"
output: html_document
---

```{r setup, include=FALSE}
## Important note — I used the sf library and patchwork library which was not a part of the data science as a field class
library(tidyverse)
library(lubridate)
library(sf)
library(patchwork)
```

For this report I will analyze historic data of shootings gathered by the New York Police Department from 2006 to 2024. The dataset includes 29744 shootings and details about the shootings like the time of day the shooting happened, location of the shooting, as well as the race, sex, and age of the perpetrator and victim.

```{r get_NYPD_data, echo=FALSE}
## Get the data from the file
url_in <- "https://data.cityofnewyork.us/api/v3/views/833y-fsy8/export.csv?accessType=DOWNLOAD"
NYPD_shootings <- read_csv(url_in)
```
I will now do a brief cleaning the data by deleting columns that aren't needed and writing the date in the right format. Below is a summary of the data columns.

```{r data_cleaning, echo=FALSE}
NYPD_shootings <- NYPD_shootings %>% select(!c(INCIDENT_KEY, JURISDICTION_CODE, LOCATION_DESC, Lon_Lat, LOC_OF_OCCUR_DESC, LOC_CLASSFCTN_DESC, X_COORD_CD, Y_COORD_CD, PERP_AGE_GROUP, PERP_SEX, PERP_RACE, VIC_AGE_GROUP, VIC_SEX, VIC_RACE)) %>% mutate(OCCUR_DATE = mdy(OCCUR_DATE), shooting = 1)
summary(NYPD_shootings)
```

I'm curious about the shootings happening each year. Is there a trend in the data. Have the shootings increased or decreased from 2006.

```{r by_year_analysis, echo=FALSE}
## Creating the shootings per year dataset
NYPD_shootings_by_year <- NYPD_shootings %>% mutate(OCCUR_DATE = year(OCCUR_DATE)) %>% summarize(.by = c(OCCUR_DATE), shooting = sum(shooting))
## Plotting the shootings per year
NYPD_shootings_by_year %>% ggplot(aes(x = OCCUR_DATE, y = shooting)) + geom_point(aes(colour = "OCCUR_DATE")) + geom_line(aes(colour = "OCCUR_DATE")) + theme(legend.position = "none", axis.text.x = element_text(angle = 90), plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5)) + labs(title = "NYPD Shootings per Year", dictionary = c(shooting = "Shootings", OCCUR_DATE = "Year"))
```

The shootings have decreased from 2006 but then jumped back up in 2020. The shootings start dropping again in 2022. Something must have happened in 2019 to 2021 to cause that huge jump. In 2020 Covid 19 pandemic was declared which may have something to do with the shootings.

I'm also curious about the shootings based on month. Do shooting totals change based on the month of the year?

```{r by_month_analysis, echo=FALSE}
## Making a dataset for the shootings that happened in each month
NYPD_shootings_by_month <- NYPD_shootings %>% mutate(OCCUR_DATE = month(OCCUR_DATE)) %>% group_by(OCCUR_DATE) %>% summarize(shooting = sum(shooting))
## Plotting the shootings that happened per month
NYPD_shootings_by_month %>% ggplot(aes(x = OCCUR_DATE, y = shooting)) + geom_point(aes(colour = "shooting")) + geom_line(aes(colour = "shooting")) + theme(legend.position = "none", legend.title.position = "top", axis.text.x = element_text(angle = 90), plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5), legend.title = element_text(hjust = 0.5)) + labs(title = "NYPD Shootings per Month", dictionary = c(shooting = "Shootings", OCCUR_DATE = "Month", race_is_same = "Race")) + scale_x_continuous(n.breaks = 7)
```

After plotting the shootings that happened in each month of the year we see most shootings happen in the middle of the year. Very little shootings happen in the beginning and end of the year.

I noticed that the graph is decreasing from the 7th month to the 12th and also from the first month to the second. If we shift the 1st month to the other side of the graph, the graph will be strictly increasing till the 7th month and then decreasing till the 1st month.

```{r month_shifted, echo=FALSE}
## Shifting the month graph to the left by taking month 1 and adding it to the end
NYPD_shootings_by_month_shifted <- NYPD_shootings_by_month %>% add_row(OCCUR_DATE = 13, shooting = 1891) %>% filter(OCCUR_DATE != 1)
## Plotting the shootings that happened per month shifted
NYPD_shootings_by_month_shifted %>% ggplot(aes(x = OCCUR_DATE, y = shooting)) + geom_point(aes(colour = "shooting")) + geom_line(aes(colour = "shooting")) + theme(legend.position = "none", legend.title.position = "top", axis.text.x = element_text(angle = 90), plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5), legend.title = element_text(hjust = 0.5)) + labs(title = "NYPD Shootings per Month", dictionary = c(shooting = "Shootings", OCCUR_DATE = "Month", race_is_same = "Race")) + scale_x_continuous(n.breaks = 7)
```

Note that the 13th month is really the 1st month of the year. Now we see that shootings increase from the 2nd to the 7th and then decrease from the 7th to the 2nd.

```{r modeling, echo=FALSE}
## Making a quadratic model of the shootings per month and plotting it
month_model_quadratic <- lm(shooting ~ OCCUR_DATE + I(OCCUR_DATE^2), data = NYPD_shootings_by_month_shifted)
## Adding the quadratic prediction onto the dataset
month_with_pred_quad <- NYPD_shootings_by_month_shifted %>% mutate(pred_quad = predict(month_model_quadratic))
## Plotting the dataset
quad <- month_with_pred_quad %>% ggplot(aes(x = OCCUR_DATE, y = shooting)) + geom_point(aes(colour = "shooting")) + geom_line(aes(colour = "shooting")) + geom_line(aes(y = pred_quad, colour = "pred")) + theme(legend.position = "bottom", legend.title.position = "left", axis.text.x = element_text(angle = 90), plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5), legend.title = element_text(hjust = 0.5)) + labs(title = "Quadratic Prediction", dictionary = c(shooting = "Shootings", OCCUR_DATE = "Month", race_is_same = "Race", colour = "", pred = "Prediction")) + scale_x_continuous(n.breaks = 7)
## Splicing the first and second halves of the month data to make two linear models
month_first <- NYPD_shootings_by_month_shifted %>% filter(OCCUR_DATE == c(2,3,4,5,6,7))
## Second half of the month data
month_second <- NYPD_shootings_by_month_shifted %>% filter(OCCUR_DATE %in% c(7,8,9,10,11,12,13))
## The two linear models for the month data
month_model_first <- lm(shooting ~ OCCUR_DATE, data = month_first)
month_model_second <- lm(shooting ~ OCCUR_DATE, data = month_second)
## Adding the predictions to the datasets
month_with_pred_first <- month_first %>% mutate(pred = predict(month_model_first))
month_with_pred_second <- month_second %>% mutate(pred = predict(month_model_second))
## Plotting the two linear predictions onto the graph
linear <- NYPD_shootings_by_month_shifted %>% ggplot(aes(x = OCCUR_DATE, y = shooting)) + geom_line(aes(y = shooting, colour = "shooting")) + geom_point(aes(y = shooting, colour = "shooting")) + geom_line(data = month_with_pred_first, aes(y = pred, colour = "pred")) + geom_line(data = month_with_pred_second, aes(y = pred, colour = "pred")) + theme(legend.position = "bottom", legend.title.position = "left", axis.text.x = element_text(angle = 90), plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5), legend.title = element_text(hjust = 0.5)) + labs(title = "Linear Prediction", dictionary = c(shooting = "Shootings", OCCUR_DATE = "Month", race_is_same = "Race", colour = "", pred = "Prediction")) + scale_x_continuous(n.breaks = 7)
## Plotting the predictions side by side
quad + linear
```

Above you can see two models of the data. The first is a quadratic model and has an ok fit. The second model is made by placing two linear models side by side and the fit is much better.

## Creating maps of the shootings

I also wanted to create some maps of the shootings because these visualizations might lead to important conclusions. I used the kjhealy New York city Borough and Precinct data sets to create the maps.

```{r maps, echo=FALSE}
## A new dataset with two columns for whether the victim survived or died to use later
NYPD_shootings_with_survive <- NYPD_shootings %>% mutate(died = ifelse(STATISTICAL_MURDER_FLAG == "TRUE", 1, 0), survived = ifelse(STATISTICAL_MURDER_FLAG == "FALSE", 1, 0))
## Only the Latitude and Longitude have NA's so the rest of the data is fine
NYPD_shootings_by_Boro <- NYPD_shootings_with_survive %>% group_by(BORO) %>% summarize(shootings = sum(shooting), died = sum(died), survived = sum(survived)) %>% ungroup()
## Using the nyc boro maps dataset from kjhealy
boro <- read_rds("https://github.com/KyNiccore/NYPD_shootings/raw/refs/heads/main/nyc_boros_sf.rds")
## Rewriting the values in the columns of the boro dataset so I can join it with the shootings data
boro <- boro %>% rename(BORO = boro_name) %>% mutate(BORO = toupper(BORO))
## Filtering out any shootings where there isn't a latitude or longitude (which only is about one hundred shootings)
NYPD_shootings <- NYPD_shootings %>% filter(!is.na(Latitude), !is.na(Longitude))
## Converting the NYPD data into a spacial object to use with geom_sf
pointssf <- st_as_sf(NYPD_shootings, coords = c("Longitude", "Latitude"), crs = 4326)
## Plotting the NYPD data on a map
ggplot() + scale_fill_brewer(palette = "Blues") + geom_sf(data = boro, aes(fill = BORO)) + geom_sf(data = pointssf, color = "red", size = 1, alpha = .25) + theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5), legend.title = element_text(hjust = 0.5)) + labs(title = "NYPD Shootings Map", dictionary = c(shootings = "Shootings", BORO = "Boroughs"))
```

In this first map we see all the shootings and where they are concentrated. It looks like the Brooklyn and Bronx Boroughs have the most shootings.

```{r join, include=FALSE}
## Joining datasets to use for the died and survived borough maps
boro <- boro %>% full_join(NYPD_shootings_by_Boro)
```
```{r bar_borough, echo=FALSE}
## adding a percent column and a column for where the label will go for the bar plot
NYPD_shootings_by_Boro <- NYPD_shootings_by_Boro %>% mutate(percent = shootings / sum(NYPD_shootings_by_Boro$shootings), y_text = if_else(BORO %in% c("MANHATTAN", "QUEENS", "STATEN ISLAND"), percent + .025, percent - .025), label = paste0(round(percent, 2), "%"))
## the x axis labels for the bar plot
x_labels <- c("BRONX", "BROOKLYN", "MANHATTAN", "QUEENS", "STATEN ISLAND")
## the bar plot with the labels
bar <- NYPD_shootings_by_Boro %>% ggplot(aes(y=percent, x=BORO, fill = BORO)) + scale_fill_brewer(palette = "Blues") + geom_bar(position="stack", stat="identity") + geom_text(aes(y = y_text, label = label), size = 3) + scale_x_discrete(labels = x_labels) + theme_classic() + theme(axis.text.x = element_text(size = 10), legend.position = "NONE")
## the blank borough map to add to the bar plot
map <- ggplot() + scale_fill_brewer(palette = "Blues") + geom_sf(data = boro, aes(fill = BORO)) + theme_void() + theme(plot.title = element_blank(), legend.position = "NONE") + labs(title = "NYPD Shootings Map", dictionary = c(shootings = "Shootings", BORO = "Boroughs"))
## plotting the bar plot with the map as an added element
bar + inset_element(map, left = .45, right = 1, top = 1, bottom = .45)
```

Here is a bar plot that shows each Boroughs shootings as percents of the total shootings. We see that 69 percent of the shootings are happening in the Brooklyn and Bronx areas. We can also see that only 0.03 percent of shootings happen in Staten Island.

I then became curious about what would happen if we looked at whether the victim died. Would the map of shootings where the victim died look the same as the map where the victim survived.

```{r map_points_died_survived, echo=FALSE}
## a dataset where the victim died
pointssf_died <- pointssf %>% filter(STATISTICAL_MURDER_FLAG == "TRUE")
## a map where the victim died
died <- ggplot() + scale_fill_brewer(palette = "Blues") + geom_sf(data = boro, aes(fill = "boro_name"), fill = "white") + geom_sf(data = pointssf_died, size = 1, alpha = .25, color = "red") + theme(plot.title = element_blank(), plot.subtitle = element_text(hjust = 0.5), legend.title = element_text(hjust = 0.5)) + labs(title = "NYPD Shootings Map", subtitle = "Shootings where the Victim Died", dictionary = c(shootings = "Shootings"))
## a dataset where the victim survived
pointssf_survived <- pointssf %>% filter(STATISTICAL_MURDER_FLAG == "FALSE")
## a map where the victim survived
survived <- ggplot() + scale_fill_brewer(palette = "Blues") + geom_sf(data = boro, aes(fill = "boro_name"), fill = "white") + geom_sf(data = pointssf_survived, size = 1, alpha = .25, color = "blue") + theme(plot.title = element_blank(), plot.subtitle = element_text(hjust = 0.5), legend.title = element_text(hjust = 0.5)) + labs(title = "NYPD Shootings Map", subtitle = "Shootings where the Victim Survived", dictionary = c(shootings = "Shootings"))
## plotting the survived and died maps side by side
survived + died
```

And . . . It does look like the graphs are the same, or at least very similar. But I want to double check this. Here are two more graphs but this time the Boroughs are colored instead of just plotting all the points. 

```{r maps_boro_died_survived, echo=FALSE}
## Plotting the NYC Boroughs and coloring based on cases where the victim died
map_died <- boro %>% ggplot(aes(fill = died)) + geom_sf() + theme_void() + theme(plot.title = element_text(hjust = 0), plot.subtitle = element_text(hjust = 0.5), legend.title = element_text(hjust = 1)) + labs(title = "Shootings Map where Victim Died", dictionary = c(shootings = "Shootings", died = "total shootings"))
## Plotting the NYC Boroughs and coloring based on cases where the victim survived
map_survived <- boro %>% ggplot(aes(fill = survived)) + geom_sf() + theme_void() + theme(plot.title = element_text(hjust = 0), plot.subtitle = element_text(hjust = 0.5), legend.title = element_text(hjust = 1)) + labs(title = "Shootings Map where Victim Survived", dictionary = c(shootings = "Shootings", survived = "total shootings"))
## Plotting the two above graphs together
map_survived + map_died
```

The graphs are so similar. Apart from the scale (there are far more shootings where the victim survived) the graphs look almost identical. There isn't any difference in the distribution of shootings by Borough when grouping the shootings by whether the victim survived.

I'm now very curious if we broke the map down by precinct since that's a variable in our data.

```{r precinct_data, echo=FALSE}
## the precinct data from kjhealy
precinct <- read_rds("https://github.com/KyNiccore/NYPD_shootings/raw/refs/heads/main/nyc_police_precincts_sf.rds")
```
```{r label_calculations, include=FALSE}
## a dataset that will be used to calculate the location of labels for our precinct map
centroids <- st_centroid(precinct, of_largest_polygon = TRUE)
## adding the locations to the previous dataset
centers <- cbind(centroids, st_coordinates(centroids))
```
```{r precinct_first_map, echo=FALSE}
## making a blank map with the labels since the map with the points will be to messy
blank_precinct <- ggplot() + scale_fill_brewer(palette = "Blues") + geom_sf(data = precinct, fill = "white") + labs(title = "NYPD Shootings Map", dictionary = c(shootings = "Shootings", BORO = "Boroughs")) + geom_sf_text(data = centers, aes(x = X, y = Y, label = precinct), size = 3, color = "black") + theme_void() + theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5), legend.title = element_text(hjust = 0.5), axis.title.x = element_blank(), axis.title.y = element_blank())
## map of the points on top of the precincts
filled_precinct <- ggplot() + scale_fill_brewer(palette = "Blues") + geom_sf(data = precinct, fill = "white") + geom_sf(data = pointssf, color = "red", size = 1, alpha = .25) + theme_void() + theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5), legend.title = element_text(hjust = 0.5), ) + labs(title = "NYPD Shootings Map", dictionary = c(shootings = "Shootings", BORO = "Boroughs"))
## plotting the two maps together
blank_precinct + filled_precinct
```

We can see all the precincts labels on the left and the points on top of the precincts on the right. To make things easier to visualize I colored the precincts based on the shootings in the next map.

```{r precinct_colored_by_shooting, echo=FALSE}
## making a new NYPD dataset grouped by precinct to use for the maps
NYPD_shootings_by_precinct <- NYPD_shootings_with_survive %>% group_by(PRECINCT) %>% summarize(shootings = sum(shooting), died = sum(died), survived = sum(survived)) %>% ungroup() %>% mutate(precinct = PRECINCT)
## joining the precinct map data with the shooting precinct data
precinct <- precinct %>% full_join(NYPD_shootings_by_precinct)
## plotting the precincts and coloring based on shootings
precinct %>% ggplot() + geom_sf(aes(fill = shootings)) + geom_sf_text(data = centers, aes(x = X, y = Y, label = precinct), size = 3, color = "white") + theme_void() + theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5), legend.title = element_text(hjust = 1)) + labs(title = "NYPD Shootings Map of Precincts", dictionary = c(shootings = "Shootings", survived = "total shootings"))
```

This map is a much nicer visualization of the number of shootings that occur in each precincts. We can see precincts 73 and 75 having the most shootings. Precinct 116 wasn't created when this shooting data was being taken. It first opened in 2024 and is the newest precinct which is why it is gray. That area used to be part of the 105th and 113th precincts.

If we look at whether the victim died or survived again will we see the same result. Will the maps look the same again?

```{r precinct_died_survived, echo=FALSE}
## map of shootings where the victim died
precinct_died <- precinct %>% ggplot() + geom_sf(aes(fill = died)) + theme_void() + theme(plot.title = element_text(hjust = 0), plot.subtitle = element_text(hjust = 0.5), legend.title = element_text(hjust = 1)) + labs(title = "Shootings Map where Victim Died", dictionary = c(shootings = "Shootings", died = "total shootings"))
## map of shootings where the victim survived
precinct_survived <- precinct %>% ggplot() + geom_sf(aes(fill = survived)) + theme_void() + theme(plot.title = element_text(hjust = 0), plot.subtitle = element_text(hjust = 0.5), legend.title = element_text(hjust = 1)) + labs(title = "Shootings Map where Victim Survived", dictionary = c(shootings = "Shootings", survived = "total shootings"))
## plotting the two maps together
precinct_survived + precinct_died
```

It happened again. The maps are nearly identical. Grouping the shootings by whether the victim died or survived didn't change which precinct the shootings happened in.

# Conclusion

* The yearly shooting graph shows a downward trend till 2020 where shootings increased drastically. They then started decreasing again in 2022. This spike I've noted may be linked to covid 19.
* The month shooting graphs showed us most shootings are happening in the middle of the year. And little shootings are happening at the beginning and ends of the year.
* The maps portion contained a couple standout conclusions
  * We saw that the Brooklyn and Bronx Boroughs have the most shootings by far and Staten Island has the fewest shootings.
  * When grouping by whether the victim died or survived the maps looked the same. The location of the shooting doesn't determine whether the shooting was fatal. The location does determine whether a shooting happens which we saw in the first map and the bar graph
  * We got to see the precincts with the most and least shootings. The cluster around 73 and 75 had a lot of shootings as well as the cluster around 44.
  * When grouping by whether the victim died or survived the same thing happened. The maps looked identical. Again we see that the location doesn't determine whether a shooting is lethal or not but location does determine whether a shooting occurs.

# Bias

There were a small number of data points missing their latitude and longitude which we couldn't use. There is also bias in these being the shootings that are documented by the police. Maybe the police is prone to document certain types of shootings or shootings in certain locations. If the police force is better at documenting in a certain area then that area might have more shootings. There also could be a large number of shootings that aren't documented. And there is also some amount of personal bias as well which I tried to mitigate by not explaining trends in the data using personal beliefs.


